---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Glamorous graphics

In a [talk at RStudio conference in 2020](https://posit.co/resources/videos/the-glamour-of-graphics/), Will Chase presented principles related to glamorous graphics. In contrast to Tufte's principles (discussed above), glamorous graphics suggest adding ink to increase the attractiveness of graphics. The ideas presented by Chase might at first seem distant from dry scientific graphics, where no other message should be conveyed than what the data shows. However, good-looking graphics might communicate your message more effectively than bad-looking graphics, even though the style itself carries some additional message. Consider the plots below. What message other than the story related to the data do they convey? 


```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-esthetic
#| fig-cap: "Different style choices also communicates messages."
#| fig-height: 10
#| fig-width: 4

library(tidyverse)
library(ggthemes)
library(ggtext)
library(cowplot)

p <- exscidata::millward |> 
  ggplot(aes(RNA, protein_synthesis, color = group)) + 
  geom_point() + 
  theme_classic() + 
  scale_x_continuous(limits = c(0, 15), expand = c(0,0)) + 
  scale_y_continuous(limits = c(0, 0.25), expand = c(0,0)) + 
  labs(x = "RNA (&mu;m mg<sup>-1</sup>)", 
       y = "Protein synthesis (Ks day<sup>-1</sup>)", 
       color = "Group") 
  
  

p1 <- p + theme_excel() + scale_color_excel() + theme(axis.title.x = element_markdown(), 
                          axis.title.y = element_markdown())


p2 <- p + theme_tufte() + theme(axis.title.x = element_markdown(), 
                          axis.title.y = element_markdown())

p3 <- p + theme_stata() + scale_color_stata() + scale_shape_stata() + 
  theme(axis.title.x = element_markdown(), 
                          axis.title.y = element_markdown())

p4 <- p + theme(axis.title.x = element_markdown(), 
                          axis.title.y = element_markdown())



plot_grid(p1, p2, p3, p4, ncol = 1)



```




## Colors 

### How to pick colors?


```{r}
#| eval: false


library(colorspace)


choose_color()


```




```{r}

peppa <- c("#6C61A6",
           "#6FB3B6", 
           "#87B9F6",
           "#007CC3",
           "#77C76E",
           "#327B47",
           "#A68979",
           "#8D7C68",
           "#FFB0DF",
           "#FE8ACE",
           "#FE9A34") 


exscidata::millward |> 
  ggplot(aes(RNA, protein_synthesis)) + 
  
  geom_point(fill = peppa[9], shape = 21, size = 6) + 
  
  theme(axis.line = element_line(color = peppa[9]), 
        axis.ticks = element_line(color = peppa[9]),
        axis.text = element_text(color = peppa[9], size = 15), 
        plot.background = element_rect(fill = peppa[4]), 
        panel.background = element_rect(fill = peppa[3]), 
        panel.grid = element_blank())
        



```





<!-- 
Why glamorous graphics? - Engage a wider audience - Aesthetics matters for understanding - "respecting the audience"

The glamour of graphics uses color, typography and layout to increase the impact of the graphic.

### Layout

#### Alignment

Top left align titles

Text should be horizontal

Axis labels should also be horizontal

Alignment should provide clean lines and symmetry

Some alignment can be left/center/right to provide overall structure to the graph.

#### Border, Grid lines, legends

Do not use borders or grid lines. If grid lines, make them lighter and more subtle.

Avoid legends, labels of the data are better or titles with color.

#### White space

Use white space to separate elements of the graph and give the reader room to understand the graph.

### Typography

#### Fonts

#### Hierarchy

This gives the reader a hierarchy of information.

### Color

What colors look good together?

```{r}

# Load required libraries
library(ggplot2)

# Create a data frame of angles and radii
angles <- seq(0, 360, by = 0.1)
radii <- seq(0.1, 1, by = 0.1)  # Vary the starting point to avoid a point at the center
df <- expand.grid(angle = angles, radius = radii)

# Create ggplot of colored points in polar coordinates
p <- ggplot(df, aes(x = angle, y = radius, color = angle, alpha = radius)) +
  geom_point(size = 8) +
  scale_color_gradientn(colors = rainbow(360)) +
  scale_alpha_continuous(range = c(1, 0.1)) +
  coord_polar(theta = "x") +
  theme_void() +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none")
  

#### Alternative

# Create a data frame of angles from 0 to 360
angles <- seq(0, 360, by = 0.2)
df <- data.frame(angle = angles)

# Create ggplot of colored points in polar coordinates
p <- ggplot(df, aes(x = angle, y = 1, color = angle)) +
  geom_point(size = 10) +
  scale_color_gradientn(colors = rainbow(360)) +
  coord_polar(theta = "x") +
  theme_void() +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none")

```

The color wheel can be used to find

-   Complementary colors that will give high contrast
-   Monochromatic will give harmonious\
-   Analogous are choosen from the same side on the color wheele colors to avoid contrast but still enable showing differences
-   Triadic colors to give more

 https://www.canva.com/colors/color-wheel/ for more on this 

#### Describing colors, Hue, Saturation and Lightness

#### Background colors

Colors can help reduce "hardness" by avoiding a white background.-->
